---
title: "trelliscopejs"
author: "Ryan Hafen"
copyright: "Ryan Hafen"
output:
  packagedocs::package_docs:
    toc: true
    toc_collapse: false
redirect: http://hafen.github.io/trelliscopejs
vignette: |
  %\VignetteIndexEntry{trelliscopejs Documentation}
  %\VignetteEngine{packagedocs::redirect}
navpills: |
  <li class="active"><a href='index.html'>Docs</a></li>
  <li><a href='rd.html'>Package Ref</a></li>
  <li><a href='https://github.com/hafen/trelliscopejs'>Github <i class='fa fa-github'></i></a></li>
---

<script src="https://unpkg.com/trelliscopejs-lib/dist/trelliscope.min.js"></script>

```{r global_options, include=FALSE}
# R output pre blocks are styled by default to indicate output
knitr::opts_chunk$set(comment = NA)

# shorthand for rd_link() - see ?packagedocs::rd_link for more information
rdl <- function(x) packagedocs::rd_link(deparse(substitute(x)))

library(rbokeh)
thm <- getOption("bokeh_theme")
thm$axis$axis_label_text_font_size <- "10pt"
options(bokeh_theme = thm)
```

# Trelliscope

Trelliscope is a visualization approach based on the idea of ["small multiples"](http://www.juiceanalytics.com/writing/better-know-visualization-small-multiples) or ["Trellis Display"](chrome-extension://oemmndcbldboiebfnladdacbdfmadadm/http://polisci.msu.edu/jacoby/uic/manuscripts/95.8.color.pdf), where data are split into groups and a plot is made for each group, with the resulting plots arranged in a grid. This approach is very simple yet is considered to be ["the best design solution for a wide range or problems in data presentation"](https://en.wikipedia.org/wiki/Small_multiple). Trelliscope makes small multiple displays come alive by providing the ability to interactively sort and filter the plots based on summary statistics computed for each group.

The **trelliscopejs** R package is an interface for creating Trelliscope displays with the goal of providing methods that are convenient to use in traditional visual / analytical workflows. For example, the package provides methods that allow you to easily generate Trelliscope displays from ggplot2 or ["tidy"](https://blog.rstudio.org/2016/09/15/tidyverse-1-0-0/) workflows.

At the core of the **trelliscopejs** is a JavaScript library, [trelliscopejs-lib](https://github.com/hafen/trelliscopejs-lib), which provides the interactive viewer for the displays, and the R package is an [htmlwidget](http://www.htmlwidgets.org/) around this library, meaning that it is easy to share your displays whether through embedding in an RMarkdown document / notebook or simply sending the resulting html file to someone or hosting on the web.

To quickly get a feel for **trelliscopejs**, here's a screen capture of it in use:

<div style="text-align: center;">
<iframe width="700" height="394" src="https://www.youtube.com/embed/u3YGLD-5R0s?rel=0" frameborder="0" allowfullscreen></iframe>
</div>

# Getting Started

## Installation

**trellscopejs** is currently available on Github. It is stable and well tested and documented and ready for CRAN, but we are making little tweaks and adding feature so frequently that it is not yet on CRAN but we plan to have it there soon. Anyway, it is ready for people to try out right now. So please do!

```r
devtools::install_github("hafen/trelliscopejs")
```

Any other package that you see in this document that is not already installed along with **trelliscopejs** will be available on CRAN and can be installed with `install.packages()`.

## facet_trelliscope

For ggplot2 users, the easiest way to get quick results with **trelliscopejs** is with the function `r rdl(facet_trelliscope())`. If you are familiar with faceting in ggplot2 using `facet_wrap()`, you can easily transform your static faceted plot into an interactive one by swapping out `facet_wrap()` with `facet_trelliscope()`.

To illustrate, let's look at the gapminder dataset available in the **gapminder** package, which provides some population statistics about many countries in the world:

```{r, message=FALSE}
library(trelliscopejs)
library(ggplot2)
library(gapminder)

glimpse(gapminder)
```

We might be interested looking at life expectancy vs. year for each country in Europe:

```{r fig.width=11.8, fig.height=10}
qplot(year, lifeExp, data = subset(gapminder, continent == "Europe")) +
  facet_wrap(~ country + continent) +
  theme_bw()
```

Now suppose we want to be able to see all countries:

```{r fig.width=11.8, fig.height=10}
qplot(year, lifeExp, data = gapminder) +
  xlim(1948, 2011) + ylim(10, 95) + theme_bw() +
  facet_wrap(~ country + continent)
```

There are too many to show on a single plot. This is where we can turn to Trelliscope.

To create the same plot with Trelliscope, swap `facet_wrap()` for the **trelliscopejs** function `facet_trelliscope()`:

```r
qplot(year, lifeExp, data = gapminder) +
  xlim(1948, 2011) + ylim(10, 95) + theme_bw() +
  facet_trelliscope(~ country + continent, nrow = 2, ncol = 7, width = 300)
```

A simple change! The output will look something like this:

<div style="margin-top:20px; margin-bottom: 20px">
<div id="ebbfa969" class="trelliscope-not-spa" style="width:850px; height:600px;"></div>
</div>

Go ahead and experiment with the interactive controls of the plot above. You can click the fullscreen button in the bottom right if you want more space. The question mark icon in the upper right corner will give you more information about how to use the viewer.

Here we used a different layout of rows and columns to illustrate Trelliscope's pagination. In the viewer you will see that there are 14 panels shown per page, and that you can use the "prev" and "next" buttons to page through the panels. You can change the layout in the "Grid" controls. Also, we can now filter the panels based on various summaries that were computed for each panel.

For example, we might want to only look at panels for a certain continent, such as Africa. Or we might want to sort the panels on mean life expectancy. You can see what variables are available opening the "Filter" or "Sort" controls.

You can see some of these controls in action in the video above, and we'll discuss more about how these summaries are computed below. If you click the question mark icon in the upper right of the viewer, you can learn more about the controls and how to use them. Feel free to explore.

There are many more options in `r rdl(facet_trelliscope())` and you can read the documentation for this function for more information and examples.

<!-- Note that Trelliscope can be useful even for trivial faceting with smaller displays. For example, suppose we want to compare the distribution of city mpg for each class:

```{r fig.width=8, fig.align="center"}
qplot(class, cty, data = mpg, geom = c("boxplot", "jitter")) +
  ylim(7, 37) + theme_bw()
```

Here, the side-by-side boxplots are essentially a trellis display in that we are looking at the same plot for each class. It might be more interesting if the boxplots were sorted by city mpg mean. While this can be done programmatically with ggplot2, we can create essentially the same plot with Trelliscope with no extra effort and interactively change how the panels are sorted as we wish.

```r
qplot(class, cty, data = mpg, geom = c("boxplot", "jitter")) +
  facet_trelliscope(~ class, ncol = 7, height = 800, width = 200,
    state = list(sort = list(sort_spec("cty_mean")))) +
  ylim(7, 37) + theme_bw()
```

In this display, the panels are by default sorted by mean city mpg, but you can open the "Sort" sidebar and rearrange the panels as you wish.

This example brings up a few more options for `r rdl(facet_trelliscope())`. First, you can specify the height and width of the panels (in pixels). While these dimensions are relative based on the size and layout of the viewer, they control the aspect ratio of the plot which is very important. Another option shown in this example is the specification of the default sort order through the `state` parameter. This is likely to become much more straightforward to specify in the future so I won't go into detail here.
 -->

## "tidy" Interface

In addition to `facet_trelliscope()`, this package comes with a more generic `trelliscope()` method that fits naturally into ["tidy"](https://blog.rstudio.org/2016/09/15/tidyverse-1-0-0/) workflows involving dplyr / tidyr / purrr / etc.

An emerging use case for analyzing data in the tidyverse is computing summaries, models, and other transformations for each group of a data set and storing the results as a data frame with one row per group, with columns of data frames containing the data for each group and possibly additional columns containing other derived data such as models, etc. A nice example of this is illustrated in "Many Models" section of [R for Data Science](http://r4ds.had.co.nz/many-models.html). That book chapter (and any necessary material preceding it) can be taken as prerequisite material to fully appreciate the visualization approach outlined in this section and fully be able to apply these techniques to your own problems. In particular, pay attention to the concepts of nested data frames and list-columns.

The example in the ["Many Models"](http://r4ds.had.co.nz/many-models.html) chapter uses the Gapminder data and illustrates grouping the data by country, nesting the data for each country into a new column of data frames, and then fitting a linear model of life expectancy vs. time for each group. In this example, to visualize the model results, the model residuals are un-nested and visualized in a single plot. This is a common visualization workflow: group the data, compute summaries or derived data for each group, pull all the results back together, and create a single visualization of the result. This is a good and useful thing to do, but additionally visualizing the detailed per-group data is invariably very enlightening. This is where Trelliscope comes in.

### Gapminder example

Let's set up a scenario in which we will produce a plot similar to the Gapminder plot we created using `facet_trelliscope()`, but using nested data frames and list-columns.

We want to plot life expectancy vs. year by country. So first we will group our data frame by country and continent and thenc call `tidyr::nest()` to obtain a data frame with one row per country and continent with each unique country's data stored in a new column of nested data frames, `data`:

```{r}
library(tidyverse)
library(gapminder)

by_country <- gapminder %>%
  group_by(country, continent) %>%
  nest()

by_country
```

We can now do all kinds of per-group operations by operating on the rows of this data frame or on elements of the columns of this data frame, and we can even attach these results as new columns of the data frame, allowing us to keep all per-group computations organized by group in a familiar data structure. For example, in "R for Data Science" takes this result and fits a model to each group using `purrr::map()`.

Let's go ahead and add the model fit as well:

```{r}
country_model <- function(df)
  lm(lifeExp ~ year, data = df)

by_country <- by_country %>%
  mutate(model = map(data, country_model))

by_country
```

We now have a new column, `model` which contains the model fit result for each country.

### Panels as list-columns

The main idea to communicate with Trelliscope in the tidyverse is that in addition to storing results of per-group summaries or models as list-columns, we also store per-group plots as list-columns and use a function `r rdl(trelliscope())` to visualize them. This is quite straightforward to do and is accomplished using a family of purrr-like `map` functions specifically for returning plot list-columns.

The functions available are:

- `map_plot()`
- `map2_plot()`
- `pmap_plot()`
- `by_row_plot()`

These are all basically wrappers around the corresponding purrr functions (`purrr::map()`, `purrr::map2()`, etc.) that indicate that an output is a list of plots. These functions can be used to operate over columns of a data frame to apply a user-defined plotting function to each group. Your plot function can return a wide variety of plot objects, including ggplot2, lattice, and htmlwidgets.

Let's look at an example. With our nested Gapminder data, let's use `map2_plot()` to add a plot column to our data set which plots both the raw data and the fitted line. `map2_plot()` allows us to specify two columns (we'll want to use the columns `data` and `model`) as input to our plot function, whereas we can only specify one column with `map_plot()`. In this example we use the [rbokeh](http://hafen.github.io/rbokeh/) package, which is an htmlwidget, to create the plots.

```{r}
library(rbokeh)
library(trelliscopejs)

country_plot <- function(data, model) {
  figure(xlim = c(1948, 2011), ylim = c(10, 95), tools = NULL) %>%
    ly_points(year, lifeExp, data = data, hover = data) %>%
    ly_abline(model)
}

by_country <- by_country %>%
  mutate(data_plot = map2_plot(data, model, country_plot))

by_country
```

The `by_country` data frame now has a new column, `data_plot` that contains the plot object for each group. In addition to storing nested data in our data frame, we are storing plots!

We can now pipe this into a call to `trelliscope()`:

```r
d %>%
  trelliscope(name = "life_expectancy")
```

The output will look like this:

<div style="margin-top:20px; margin-bottom: 20px">
<div id="85d4763" class="trelliscope-not-spa" style="width:850px; height:600px;"></div>
</div>

The `r rdl(trelliscope())` function takes a data frame as input. At a minimum, it expects this data frame to have a plot column (if there is more than one plot column, you can specify which one to use with the `panel_col` argument). Any additional atomic columns are made available as variables with which you can navigate the panels in your Trelliscope display.

The additional arguments to `r rdl(trelliscope())` are very similar to that of `r rdl(facet_trelliscope())`. All **trelliscopejs** displays must have names. In the ggplot2 examples above, if a name is not specified, it is inferred from the ggplot2 object.

Looking at this example, this is a conceptually very interesting way to think about what we are doing with faceted plots. As noted, we are creating plots as per-group "summaries", and using other summary statistics to help us navigate the space of these plots. *"Summary plots"* based on summary statistics coupled with detailed *"plot summaries"* of the raw data from which summaries are computed is a powerful combination for exploratory data analysis.

Although the `facet_trelliscope()` approach with ggplot2 is usually more concise and elegant, I tend to prefer this tidy approach. First, there is more flexibility in what plotting library you use (you can still use ggplot2, of course). But I prefer the flexibility of being able to shape the data for each subset in the way I want for plotting (group first and reshape in an endless number of ways rather than specify the grouping with the "facet" functions).

### Control Over Cognostics

Summaries that are used to navigate a Trelliscope display have a special name, *"cognostics"*. This is a term coined by John Tukey that you can read more about [here](https://projecteuclid.org/download/pdf_1/euclid.aos/1043351250).

**trelliscopejs** tries to make the specification of cognostics as easy as possible. As you have noticed in tidy example above, any available atomic variables in a data frame are automatically turned into cognostics. Also, note that since the original data is preserved in the nested `data` column, `trelliscope()` detects this and computes automatic cognostics for the variables in that column. Currently these automatic summaries are simply means if the variable is numeric, and the value of any categorical variable that is fixed within each group. The same is done for `facet_trelliscope()` as you may have noticed above. We will be adding more automatic cognostics in the future, particularly geared toward extracting cognostics based on the context of what is being plotted (see "upcoming work" below).

The **trelliscopejs** package has some convenience functions for specifying cognostics that help describe their behavior in the Trelliscope viewer. One of the most useful attributes is a cognostics *description* that helps the user navigating the displays know more about what the summary statistics that they are looking at mean.

Let's add a new cognostic to our gapminder data frame that illustrates the use of `cog()`. Suppose we wish to be able to organize the panels of the display with respect to a goodness-of-fit measure of our linear model. Let's add a new column, `resid_mad` that computes the median absolute deviation of the residuals for each country's fit.

```{r}
by_country <- by_country %>%
  mutate(resid_mad = cog(
    map_dbl(model, ~ mad(resid(.x))),
    desc = "median absolute deviation of residuals"))
```

The variable `resid_mad` will now be available in a new Trelliscope display we create with this data, and the description "median absolute deviation of residuals" will be used in the viewer to help users be oriented with what this variable means. Note that we could have left out the `cog()` function and the variable would still be available in the display, just without the description.

You can look at the documentation for `r rdl(cog())` to see what other options are available.

### Customizing displays with data frame operations

Note that we can use data frame operations such as dplyr's `arrange()` and `filter()` prior to calling `trelliscope()` to customize Trelliscope displays.

For example, I might want the default sort order of panels to be by `resid_mad` from high to low. I can simply order the data frame and pass the result into `trelliscope()` to achieve this effect:

```r
by_country %>%
  arrange(-resid_mad) %>%
  trelliscope(name = "by_country_lm", nrow = 2, ncol = 4)
```

Or suppose I only want to make a display for the country of Africa. I can use `filter()` to keep records only for Africa and pass this result to `trelliscope()`:

```r
by_country %>%
  filter(continent == "Africa") %>%
  trelliscope(name = "by_country_africa_lm", nrow = 2, ncol = 4)
```

## Collections of Displays

While the examples shown so far create one-off displays, it is possible to build and maintain a collection of Trelliscope displays that are bundled together in the same viewer. This is useful for tracking and sharing multiple displays in once place, and for future functionality which will allow the ability to show panels from multiple displays created from the same grouping side by side.

To achieve this, you can simply specify a `path` pointing to a directory where you would like multiple displays to be organized, as an argument to `facet_trelliscope(..., path = "__my_path__")` or `trelliscope(..., path = "__my_path__")`. Any display created will be placed inside this directory and an inventory of displays will be maintained and will be available to select from in the viewer. Note that you can also use the `group` argument to help organize multiple displays into groups.

Using the `path` argument even when not creating a collection of displays is a good idea when you want your display to persist or if you want to control wher it is so you can share it.

## Modifying Displays

Functions will be added in the future that allow you to treat a display as an object and modify its properties (state, name, cognostics, location, etc.) without re-creating the display every time.

## Embedding / sharing

While all of the examples so far have created a single application that fills the window of your viewer pane or web browser, you can embed multiple trelliscope displays inside of RMarkdown documents or RMarkdown Notebooks. When using **trelliscopejs** in RMarkdown documents, each display is embedded within the flow of the document (and dimensions can be controlled in the usual way), but an additional "fullscreen" button appears in the bottom right corner allowing toggling between full window mode.  In RMarkdown Notebooks, however, you do not get the fullscreen option because Notebooks embed htmlwidgets as iframes rather than as an integrated part of the page.

Note that if you want to embed a plot in an RMarkdown notebook, you must specify `self_contained = TRUE` in your call to `facet_trelliscope()` or `trelliscope()`. This causes all of the plot dependencies to be embedded in an html page instead of saved as separate files and then loaded as the viewer needs them. This will hopefully change in the future, but `self_contained = TRUE` should be avoided whenever possible because especially when the number of panels is large, it's much more efficient to save the panels as separate files and load them on demand rather than embed them all in the page and wait for them to load at the beginning.

**trelliscopejs** displays are also easy to share as standalone apps on the web through Github or a simple web server. For example, some of the examples from this document have been made available through Github (see [here](http://hafen.github.io/trelliscopejs-demo/housing/), [here](http://hafen.github.io/trelliscopejs-demo/gapminder/), and [here](http://hafen.github.io/trelliscopejs-demo/mpg/)).

## Low-Level Functions

**trelliscopejs** provides several low-level functions useful for building new interfaces. I will add more documentation for these in the future, but they are used as the building blocks for `facet_trelliscope()` and `trelliscope()` and examples of their usage can be found in the [source code for these functions](https://github.com/hafen/trelliscopejs/blob/master/R/trelliscope.R).

# Examples

Here are a few more interesting examples of Trelliscope. You can check out my [blog](http://ryanhafen.com/blog/) for upcoming posts on other interesting displays.

Here is the resulting display:

<div style="margin-top:20px; margin-bottom: 20px">
<div id="32b2482a" class="trelliscope-not-spa" style="width:850px; height:600px;"></div>
</div>

If you'd like to interact with this display, try sorting primarily on life expectancy and /or filtering by continent to start, and see if you can find anything interesting.

The data being plotted is fairly simple -- we see a usually steady increase in life expectancy over time, with varying mean by country. However, the eye is able to quickly catch deviations from the normal pattern such as the dips in Rwanda and Cambodia or the peaks in life expectancy in several African countries in the late 80s / early 90s with life expectancy decreasing after. 142 countries is almost small enough that you might stick all of them on one static trellis display, but the added functionality of filtering and sorting helps navigate the displays in an interactive query-driven manner.

<!-- small multiples vs. animation.... While animation can make general impressions about things going on in the data, it is very difficult to quantitatively capture what is really going on over time because you can't compare one frame to another. -->

## Housing data

While the power of small multiples often lies in making rapid comparisons across many panels, it can also be useful to study individual subsets of data over a large space of subsets. A good example of this is with data available in the ["housingData"](https://github.com/hafen/housingData) package.

This package has a dataset, "housing" that gives the monthly median list and sold price for residential homes by US county, provided by Zillow. Let's take a look at the median list price over time by county.

Here, we are going to compute our own custom summary "cognostics", including the slope of a fitted line, the median list and sold price, the number of non-NA observations, and a special "href" cognostic, which will provide a link to Zillow showing the homes for sale in the county.

For this one I would recommend using rbokeh over ggplot2 simply because of the time it takes to generate the panels with ggplot2.

```r
library(housingData)
housing$state <- as.character(housing$state)
housing$county <- as.character(housing$county)

slope_fn <- function(x, y)
  coef(lm(y ~ x))[2]

d <- housing %>%
  group_by(county, state) %>%
  nest()

d <- d %>%
  mutate(
    cogs = map(data, function(x, state, county) {
      data_frame(
        slope = slope_fn(x$time, x$medListPriceSqft),
        mean_list = mean(x$medListPriceSqft, na.rm = TRUE),
        mean_sold = mean(x$medSoldPriceSqft, na.rm = TRUE),
        n_obs = length(which(!is.na(x$medListPriceSqft)))
      )
    }),
    zillow_link = sprintf("http://www.zillow.com/homes/%s_rb/",
      gsub(" ", "-", paste(county, state)))
  ) %>%
  unnest(cogs) %>%
  filter(n_obs > 1)

d <- d %>%
  mutate(
    panel = map_plot(data, function(x) {
      figure(xlab = "time", ylab = "median list price / sq ft", toolbar = NULL) %>%
        ly_points(time, medListPriceSqft, data = x, hover = x)
    })
  )

d %>%
  trelliscope(name = "list_vs_time",
    desc = "monthly median list price vs. time for 2984 US counties from 2008–2016")







library(housingData)

lm_coefs <- function(x, y)
  coef(lm(y ~ x))

d <- housing %>%
  group_by(county, state) %>%
  summarise(
    slope = cog(lm_coefs(time, medListPriceSqft)[2], desc = "list price slope"),
    mean_list = cog(mean(medListPriceSqft, na.rm = TRUE), desc = "mean list price / sq ft"),
    mean_sold = cog(mean(medSoldPriceSqft, na.rm = TRUE), desc = "mean sold price / sq ft"),
    n_obs = cog(length(which(!is.na(medListPriceSqft))),
       desc = "number of non-NA list prices"),
    zillow_link = cog_href(
      sprintf("http://www.zillow.com/homes/%s_rb/",
        gsub(" ", "-", paste(county, state)))[1],
      desc = "zillow link"),
    panel = panel(
      figure(xlab = "time", ylab = "median list price / sq ft", toolbar = NULL) %>%
        ly_points(time, medListPriceSqft,
          hover = data_frame(time = time, mean_list = medListPriceSqft)))
  ) %>%
  filter(n_obs > 1) %>%
  trelliscope(
    name = "list_vs_time",
    desc = "monthly mean list price vs. time for 2984 US counties from 2008–2016",
    state = list(labels = c("county", "state")))
```

Here is the resulting display:

<div style="margin-top:20px; margin-bottom: 20px">
<div id="8a43f2dd" class="trelliscope-not-spa" style="width:850px; height:600px;"></div>
</div>

Here we used the default of a single panel per page. While we can increase the grid size and make comparisons across counties, we are mainly interested in what is going on within each county. Because of this, we also did not fix the axes of every panel to have the same limits. Generally, you will want to do this as it makes comparisons across panels much more meaningful.

By simply paging through the panels, which by default are not very meaningfully sorted alphabetically by county and then state, we begin to get a feel for the data. The first 3 panels exhibit quite a bit of variation, while the fourth, Ada County, ID appears to be much more smooth, with a pronounced drop and rise in home prices over the course of the decade. We will see this kind of pattern in many other counties. Simply mindlessly paging has already given us some insights. Continuing on, we see Adams County, CO exhibiting a slight decrease in list price followed by a sharp incline from 2012 onward in what looks like about a 50% jump over 4 years. Certainly not all counties are the same when it comes to their home price history or whether or not they were effected by the housing crisis.

If you keep looking through you will see some counties with truly bizarre patterns that in some cases call data integrity into question (notice how several counties have a severe change point between July and August 2011, for example).

It is also fun to use this display to investigate the trend of home prices around where you live. You can do a regex filter on your county's name or simply filter on your state and look at all the counties arranged by price or slope.

Another interesting way to look at the dislays is sorted on the slope, while filtering out counties that have a small number of observations. In which counties is real estate booming / busting?

This display works well as an app for consumers to play with to explore home prices, and it was nearly effortless to create it. However, if we are doing modeling or analysis of this data, exploring the detail in this simple display gives us plenty to think about in terms of what we might do next.

# Why

## Why Small Multiples

The power of small multiples has hopefully been evident through some of the examples shown so far. However, it is worthwhile talking about why they are useful so that you know when you might want to use them.

### Simple to make

Creating small multiple plots, particularly with trellicopejs, is as simple as specifying how to break up the data and what to plot for each subset. It's easy to do and it's quick, which are very attractive qualities for exploratory data analysis, where not every idea sticks and you need to be able to rapidly iterate.

### Efficient to consume

"The reader can quickly learn to read an individual chart and apply this knowledge as they scan the rest of the charts. This shifts the reader's effort from understanding how the chart works to what the data says. That’s a worthy goal in all data presentation.
Small multiples enable comparison across variables and reveal the range of potential patterns in the charts." ([Zach Gemigagni](http://www.juiceanalytics.com/writing/better-know-visualization-small-multiples)).

"Small multiples are economical: once viewers understand the design of one slice, they have immediate access to the data in all the other slices... as the eye moves from one slice to the next, the consistency of the design allows viewers to focus on changes in the data rather than on changes in graphical design." ([Edward Tufte, Visual Display of Quantitative Information](https://www.edwardtufte.com/tufte/books_vdqi)).

### Harness the human brain

The human brain is remarkably adept at finding patterns, and having many examples to look at simultaneously helps differences stand out even more.

A [ProPublic article on small multiples](https://www.propublica.org/nerds/item/a-big-article-about-wee-things) delves a bit more into why this is the case:

"The reason we can make comparisons so easily is because these small multiples take advantage of the built-in capabilities of our visual system. Specifically, something called [preattentive processing](https://en.wikipedia.org/wiki/Pre-attentive_processing) -- cognitive operations that can be performed prior to focusing attention on any particular region of an image... Our brains aren't like scanners, they throw away most of what the eyes see. But they are good at perceiving simple visual features like color or shape or size, and in fact they do it amazingly fast without any conscious effort. That means we can tell immediately when a part of an image is different from the rest, without really having to focus on it. Our minds are really good at spotting one or two differences when everything else is the same."

Even when not looking for anomalous or "strange" behavior in data, small multiple plots can do a good job of revealing subtle features of a dataset.

### Avoids overplotting and allows more detailed views

Even if we are not using small multiples to find visual features that stand out, they are useful because they help eliminate overplotting. A lot of data in a small space can be confusing for visualization consumers, and small multiples break that data up into pieces that are displayed independently in a grid, allowing for greater detail of the data being plotted in each grid cell, particularly when these cells can be paginated and displayed at a higher resolution.

The ability to slice data into many plots also helps deal with higher dimensional data. Small multiples handle higher dimensions by using one or more of the variables to determine how to break the data up, and then for each subset plotting the data for one or two variables.  In most cases this is *much* more effective than 3d visualization, and can scale to higher dimensions than 3. One needs to look no further than `library(lattice); example(xyplot)` to see this idea in action.

<!--
For example:

```r
xyplot(NOx ~ C | EE, data = ethanol,
  prepanel = function(x, y) prepanel.loess(x, y, span = 1),
  xlab = "Compression Ratio", ylab = "NOx (micrograms/J)",
  panel = function(x, y) {
    panel.grid(h = -1, v = 2)
    panel.xyplot(x, y)
    panel.loess(x, y, span=1)
  },
  aspect = "xy")
```
-->

### Experience

If it counts as a final argument, experience has shown small multiples to be extremely useful in practice. This is not true just for me, but for several visualization experts.

For example, Edward Tufte said "for a wide range of problems in data presentation, small multiples are the best design solution."

Another expert to look to is Bill Cleveland, who is one of the minds behind Trellis Display and who uses it extensively to this day in many noteworthy analyses.

Not that my opinion counts nearly as much as these two, but on a personal note, I learned to appreciate the power of small multiples while I was a student of Bill Cleveland. While visualization was not the main topic of research in my graduate studies, in all of the applied work I did during that time (and since), small multiples have been at the forefront of nearly every interesting discovery.

<!--
While it's not as pretty as crazy interactive visualizations with buttons, knobs, brushing, etc. it's very simple to create and very effective.
-->

## Why Trelliscope

We are trying to design **trelliscopejs** in such a way that if you are already sold on using small multiples, you should see no reason why to not use **trelliscopejs** to generate Trelliscope small multiple displays.

Keep in mind that the main addition that Trelliscope brings to small multiples is the ability to interactively sort and filter panels of a small multiples display based on "cognostic" summaries of the data.

### Easy to specify

Small multiple plots are already easy to create in R with packages like ggplot2 or lattice. As we've seen in examples above, it is not really any more work to create a small multiples plot with **trelliscopejs**, particularly when cognostics are automatically computed for you.

### Interactivity for free

A really nice feature of Trelliscope is that you get interactivity for free. In statistical graphics, particularly exploratory data analysis, interactivity is sometimes difficult to achieve because it usually requires a lot more time and work than simple static graphics. The audience for the visualizations is usually ourselves, and we are in a hurry to see what's in the data or how well our model is working, and are not willing to put in hours or days of time on some interactive feature when we know we may discard the plot quickly. Often our plots are "interactive" in that we can rapidly create new static plots faster than we can come up with any interaction to do so for us. But if interactivity is easy and comes for free, there's no reason not to use it.

A nice feature of Trelliscope interactivity is that the interaction paradigm is the same for every display. Instead of creating many different apps that users might need to figure out how to use independently, once they learn how to use one Trellscope display, they are set (see "displays as apps" section below).

<!-- The simple interaction paradigm is also useful for provenance - the sorting and filtering of the panels is transient. -->

### Scalable

Perhaps one of the most unrealized features of small multiples, and one we are trying to tackle with Trelliscope, is the natural fit small multiple plots have with big data. Often the approach to visualizing a large data set is computing summaries and making a single visualization based on the summaries. As we've mentioned before, you can lose a lot of important information that way. A perfect companion to plots of summaries is small multiple plots which can be navigated with respect to the summaries, and provide much-needed detail into the data, regardless of size.

Small multiples with Trelliscope scales because we never *need* to be able to see all the panels, we just need to be *able* to, and we need to be guided to those panels that have the most interesting content. Cognostics help us answer that. We have used the previous incarnation of Trelliscope to flexibly visualize the raw data of hundreds of gigabytes of data on a Hadoop cluster (see [here](https://www.youtube.com/watch?v=KatkJx9Ui0o) for an example).

<!-- One of the reasons multi-panel displays are effective is that they allow you to make visual comparisons across groups. Often the number of groups is so large that it is difficult to look at all the panels at once, and in Trelliscope we can -->

### Trelliscope displays as "apps"

While the Trelliscope viewer is meant to be used primarly in the exporatory data analysis process, we have found that in many cases these displays are also useful and insightful for non-data-scientists to consume (take the housing example above, for example). In fact, in my previous job we used this approach frequently with domain scientists. The displays are easy enough for any data scientist to create without needing to know how to write an interface or an app, while the resulting displays are intuitive enough for domain experts to navigate, with the interactivity allowing them to feel and learn from their data. These displays provided an excellent communication medium when sitting down with a domain expert to discuss the data.

In addition to exploratory displays of data as apps, Trelliscope displays can also act as complete replacements of more traditional statistical apps. Many statistical apps I have seen consist of a few inputs the user can specify and a plot that updates as these inputs change. It is possible cast most such cases as a Trelliscope app, with the full space of input combinations being represented as cognostics and the output being the panel of a Trelliscope display representing the output for each input combination. An additional benefit of this approach is that many outcomes can be compared simultaneously.

# trelliscopejs-lib

**trelliscopejs** is an [htmlwidget](http://www.htmlwidgets.org/) interface to the [trelliscopejs-lib](https://www.npmjs.com/package/trelliscopejs-lib) JavaScript library. While this is a (probably) rare example of an htmlwidget where both the JavaScript and R libraries are entirely written and maintained by the same entity, trelliscopejs-lib is designed to be language-agnostic, meaning that for a new language to provide a binding to trelliscopejs-lib, it simply needs to provide a JSON specification that trelliscopejs-lib knows how to handle. The schema for this spec is settling in, but is still subject to change. More documentation will be provided as demand necessitates.

For those who are interested, trelliscopejs-lib is written using [React](https://facebook.github.io/react/), [redux](http://redux.js.org/), and [Webpack](https://webpack.github.io/). Other major components include [material-ui](http://www.material-ui.com/#/) and [crossfilter](https://www.npmjs.com/package/crossfilter2), and [d3](https://d3js.org/).

The library is available on [npm](https://www.npmjs.com/package/trelliscopejs-lib) and therefore over CDN using [unpkg](http://unpkg.com/). All dependencies, fonts, css, etc. are bundled into the library and it weighs in at a little over 1MB.

### Embedding trelliscopejs apps

It is easy to embed the library and corresponding Trelliscope displays into another web page or web application.

NOTE: this is not required knowledge for anyone who is simply creating displays with the R package, but is provided as a reference for others who may want to take displays that have been deployed and embed the viewer in their own web applications.

First, you need to load the trelliscopejs-lib library with this script tag (check for the latest version):

```html
<script src="https://unpkg.com/trelliscopejs-lib/dist/trelliscope.min.js"></script>
```

This attaches a function to the window called `trellscopeApp()` that is used to instantiate an app in a div.

Then, you create a div where you would like to embed the app. The div needs to have an ID that matches the ID provided for a given display. For example, the housing example above is shared in a github repository and in the `appfiles` subdirectory of the app, there is a text file `id`: https://github.com/hafen/trelliscopejs-demo/blob/gh-pages/housing/appfiles/id.

```html
<div id="8a43f2dd" class="trelliscope-not-spa" style="width:800px; height:500px;"></div>
```

We give this div a class "trelliscope-not-spa" to help the application know that we are embedding the display and therefore it is not a "single-page-application". You can embed it as a full-page application if you would like and would omit the "class" and "style".

Finally, at the end of your html file (or wherever it is in your app that you want to make the app appear), you need to point `trelliscopeApp()` to the id of the div it should populate and your app's config.jsonp that tells the viewer where all the components are. In the case of this example, we can point directly to the housing app files as they are hosted on github pages:

```html
<script>
  (function() {
    trelliscopeApp('8a43f2dd',
      'http://hafen.github.io/trelliscopejs-demo/housing/appfiles/config.jsonp');
  })();
</script>
```

In the future I envision `trelliscopeApp()` passing back "get" and "set" methods so that other applications can drive the behavior of a **trelliscopejs** app.

# History

I've been working on ideas related to Trelliscope since I was a graduate student less than a decade ago. Web technology has come a long way since then and Trelliscope has gone through several iterations ranging from ad hoc scripts to an implementation that scales to run on Hadoop clusters.

The most recent previous incarnation is the ["trelliscope"](http://deltarho.org/docs-trelliscope/) package, part of the [DeltaRho](http://deltarho.org/) project. **trelliscopejs** is designed to be the successor of this package and it is hoped that those using this package are able to migrate their workflows to **trelliscopejs**. There are some use cases that **trelliscopejs** can't yet handle that the old trelliscope package can, such as integration with big data systems like Hadoop and the ability to generate panels on the fly, but supporting these options is in the road map for **trelliscopejs**.

The old trelliscope package does have a very experimental conversion function that will help users convert their old displays to be compatible with **trelliscopejs**:

```r
library(trelliscope)
vdbConn("_path_to_vdb_")
vdbConvert()
```

# Upcoming Work

What's available now represents quite a lot of work, but there is much more planned for **trelliscopejs**.

### Automatic cognostics

The biggest thing on the horizon and the one I'm most excited about is the idea of "automatic cognostics". The idea is to automatically derive metrics that will be useful for navigating the panels based on the context of what is being plotted. For example, if plotting a time series, there is a useful class of metrics that can capture time-series-related plots. If plotting a scatterplot, there's [scagnostics](https://research.tableau.com/sites/default/files/Wilkinson_Infovis-05.pdf), among other things. Using plotting systems like ggplot2 or rbokeh, the plot objects can be analyzed to derive context. Barret Schloerke, a graduate student at Purdue, is working on these ideas right now.

### Axes almost always matter

If you haven't noticed, axis scales are important when viewing a small multiple display. As stated [here](http://www.juiceanalytics.com/writing/better-know-visualization-small-multiples), "Small multiples should share the same measures, scales, size, and shape. Changing one of these factors undermines ability for people to re-use their understanding of the chart."

Currently you have to manually specify axis limits with **trelliscopejs**, but we plan to soon add the ability to specify your axes as "same", "sliced", or "free".

Also, axes take up a lot of screen real estate, and when all the axes are the same, we can save that space and open up more space for visual comparison by moving the axes to the margins of the grid. This will require the ability to decompose panels into content and axes and will take some work, but is an important priority.

### Grid placement sometimes matters

Sometimes the exact grid placement of the panels makes a big difference in the effectiveness of a small multiples display. As stated [here](http://www.juiceanalytics.com/writing/better-know-visualization-small-multiples), "Placement of the small multiples charts should reflect some logical order, e.g. dimensional matrix, geography, or time. This helps the user quickly find the charts that are interesting to them."

Currently placement in **trelliscopejs** is just a sequential grid dictated by the sort order of the cognostics. We plan to investigate other ways to allow users to specify the layout of the grid such as each page.

### Scalability

In the previous implementation of Trelliscope, our initial focus was on very large datasets. This was out of necessity because of some large datasets we were working with, but also because these implementations have been products of funding from the DARPA XDATA "big data" initiative. I have since found that the vast majority of use cases for Trelliscope involve small to moderate-size data. Because of this, with **trellscopejs**, I focused first on the simplicity of the interface, and we will then work toward scalability.

The previous Trellicope implementation was based on [datadr](http://deltarho.org/docs-datadr/), a back-end agnostic framework for general-purpose distributed computing, and ran on things like Hadoop with a MapReduce-like interface. I have strong hopes for systems like sparklyr to eventually support the kind of dplyr-like operations that the **trelliscopejs** interface requires (remote arbitrary R code execution on subsets and support for list-columns) such that **trelliscopejs** simply plugs in and immediately becomes a very easy-to-use big data visualization platform.

### Viewer improvements and more modes of interaction

There are many more modes of interaction we would like to build in to the Trelliscope viewer. For example, date/time cognostics should have their own special mode of interaction. It would also be able to filter based in views of combinations of cognostics. The new React framework we are using for the JavaScript viewer facilitates rapid development and stability of such features.

We would also like to make the viewer be able to track and share state, so an analyst can share interesting discoveries with others. This should also be straightforward.

We also plan to make the viewer support more ways to serve panels of a display. Currently, **trelliscopejs** pre-generates panels as json files and the viewer loads them on demand. It would be good to have more options for requesting panels, and this will particularly be useful when working on scaling up to very large data sets.

### Other languages?

It would be really cool to see other languages become trelliscopejs-lib interfaces but we're not in the position to work on that.

<script>
  (function() {
    trelliscopeApp('ebbfa969', 'https://hafen.github.io/trelliscopejs-demo/mpg/appfiles/config.jsonp');
    trelliscopeApp('32b2482a', 'https://hafen.github.io/trelliscopejs-demo/gapminder/appfiles/config.jsonp');
    trelliscopeApp('8a43f2dd', 'https://hafen.github.io/trelliscopejs-demo/housing/appfiles/config.jsonp');
    trelliscopeApp('85d4763', 'https://hafen.github.io/trelliscopejs-demo/gap1/appfiles/config.jsonp');
  })();
</script>
